---
title: "Dependency heaviness analysis for package cola"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
---

```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    warning = FALSE,
    fig.align = "center")
```


[*cola*](https://bioconductor.org/packages/cola/) is a package for consensus partitioning
analysis. It implements comprehensive functionalities not only for consensus partitioning, but also
for downstream anlaysis such as dimension reduction analysis, signature analysis and functional
enrichment analysis. It integrates a lot of other packages. The dependency heaviness analysis
is as follows:


```{r, eval = FALSE}
library(pkgndep)
x = pkgndep("cola")
```


```{r, echo = FALSE}
library(pkgndep)
lt = load_all_pkg_dep()
x = lt[["cola"]]
pdf(NULL)
size = plot(x, help = FALSE)
invisible(dev.off())
```


```{r, fig.width = 36.2, fig.height = 10.96}
plot(x)
```

You can drag the plot into a new tab if it is too small to read. The
dependency analysis for *cola* is also available at
https://pkgndep.github.io/prefix_c/cola_dependency_report.html.


The dependency heaviness analysis shows the number of total dependency
for *cola* is 248, which means, if the full functionality of *cola* is required
by a user, he or she needs to install all 248 upstream packages. *cola*
performs consensus partitioning as its core analysis which is expected to be
very frequently used by users, while other downstream analysis such as
functional enrichment analysis are less used. On the other hand, dependency packages
for downstream analysis contribute very high heaviness to *cola*. For example,
package *clusterPrifiler* which is for functional enrichment analysis
contribute a heaviness of 91 and package *ReactomePA* which provides Reactome
pathways for enrichment analysis contribute a heaviness of 96. Since we are
also the developers of package *cola*, we arranged the parents of *cola* in a
way that only package related to the core analysis were put as strong parents,
while those for secondary analysis were put as weak parents. This makes the
number of strong dependencies of *cola* reduced to only 64.



<br>
<br>
<br>
<br>
<br>
<br>
<br>

<script>
$(function() {
  $("img").first().css("max-width", "150%");
});
</script>

